Nonparametric Noise models for the Gaussian Process

### 2012-08-20 notes/paper.tex
More notes and a bit of derivation on resampling noise parameters.

Imagine we have a time series with observations $y_t$ at times $x_t$ with $t \in [0,T]$. We model this data by assuming a latent Gaussian process
\begin{equation}
\bm{f} \sim \GP{0}{\bm{K}(\bm{x},\bm{x})}
\end{equation}
We model the noise as a non-parametric mixture model using the Dirichlet process. Let
\begin{equation}
\bm{G} \sim \DP{\alpha}{H}
\end{equation}
be a Dirichlet process with concentration parameter $\alpha$ and base measure $H$. For each time $t$ we introduce a noise variable $\epsilon_t \sim G$. We then model the observation $y_t = \bm{f}(x_t) + \epsilon_t$.

@@

We can perform inference in this model using a collapsed Gibbs sampler. In order to work with the CRP representation of the Dirichlet process we introduce a new variable $z_t$ which will represent CRP partition that datapoint $t$ belongs to. For each CRP partition $n$ we represent the cluster parameters using $\theta_n$.

@@

\paragraph{Sampling $z_t$} The conditional distribution of $z_t$ can be written as follows
\begin{eqnarray*}
p(z_t | \bm{K}, \bm{y}, \bm{z}_{\lnot t}, \bm{\theta}) & \propto & \int p(\bm{y} | \bm{f}, \bm{z}_{\lnot t}, z_t, \bm{\theta} ) p(\bm{f} | \bm{K}) d \bm{f}, \\
 & = & \int p(y_t | f_t, \theta_{z_t} ) p(f_t | \bm{y}, \bm{z}_{\lnot t}, \bm{\theta}, \bm{K}) df_t.
\end{eqnarray*}

The key bit is that $p(f_t | \bm{y}, \bm{z}_{\lnot t}, \bm{\theta}, \bm{K})$ is a Gaussian process where every datapoint contributes noise that is dependent on the CRP partition it belongs to.

\begin{equation}
\begin{bmatrix}
\bm{y} \\ f_t
\end{bmatrix}
\sim \mathcal{N}\left(\bm{0},
\begin{bmatrix}
\bm{K}(\bm{x}, \bm{x}) + \bm{\Sigma}& \bm{K}(\bm{x},\bm{x_t}) \\
\bm{K}(\bm{x_t}, \bm{x}) & \bm{K}(\bm{x_t},\bm{x_t})
\end{bmatrix}
\right)
\end{equation}

@@

We know that \cite{gpml} $p(\bm{f} | \bm{K})$ can be represented as a multivariate Gaussian distribution.

%  & = & \int p(y_t | f_t, \theta_{z_t} ) p(f_t | \bm{f}, \bm{y}, \bm{z}_{\lnot t}, z_t, \bm{\theta}) p(\bm{y}_{\lnot t} | \bm{f)_{\lnot t},  \bm{z}_{\lnot t}, \bm{\theta}) p(\bm{f} | \bm{K}) d \bm{f}, \\

@@

>>>>

Imagine we have a time series with observations $y_t$ at times $x_t$ with $t \in [0,T]$. We model this data by assuming a latent Gaussian process
\begin{equation}
\bm{f} \sim \GP{0}{\bm{K}(\bm{x},\bm{x})}
\end{equation}
We model the noise as a non-parametric mixture model using the Dirichlet process. Let
\begin{equation}
\bm{G} \sim \DP{\alpha}{H}
\end{equation}
be a Dirichlet process with concentration parameter $\alpha$ and base measure $H$. For each time $t$ we introduce a noise variable $\epsilon_t \sim G$. We then model the observation $y_t = \bm{f}(x_t) + \epsilon_t$. In this work we restrict ourselves to the case where $H$ is a Normal-Inverse Gamma distribution to parameterize the mean and variance of a normal mixture component.

@@

We can perform inference in this model using a collapsed Gibbs sampler. In order to work with the CRP representation of the Dirichlet process we introduce a new variable $z_t$ which will represent CRP partition that datapoint $t$ belongs to. For each CRP partition $n$ we represent the cluster parameters using $(\mu_n, \sigma^2_n) = \theta_n$.

@@

\paragraph{Sampling $z_t$} The conditional distribution of $z_t$ can be written as follows
\begin{eqnarray}
p(z_t | \bm{K}, \bm{y}, \bm{z}_{\lnot t}, \bm{\theta}) & \propto & \int p(\bm{y} | \bm{f}, \bm{z}_{\lnot t}, z_t, \bm{\theta} ) p(\bm{f} | \bm{K}) d \bm{f}, \\
 & = & \int p(y_t | f_t, \theta_{z_t} ) p(f_t | \bm{y}_{\lnot t}, \bm{z}_{\lnot t}, \bm{\theta}, \bm{K}) df_t.
\end{eqnarray}

The key bit is that $p(f_t | \bm{y}, \bm{z}_{\lnot t}, \bm{\theta}, \bm{K})$ corresponds to the prediction of a Gaussian process where every datapoint contributes noise that is dependent on the CRP partition it belongs to.

\begin{equation}
\begin{bmatrix}
\bm{y} \\ f_t
\end{bmatrix}
\sim \mathcal{N}\left(\bm{0},
\begin{bmatrix}
\bm{K}(\bm{x}, \bm{x}) + \bm{\Sigma} & \bm{K}(\bm{x},\bm{x_t}) \\
\bm{K}(\bm{x_t}, \bm{x}) & \bm{K}(\bm{x_t},\bm{x_t})
\end{bmatrix}
\right)
\end{equation}

@@

We know that \cite{gpml} $p(f_t | \bm{y}, \bm{z}_{\lnot t}, \bm{\theta}, \bm{K})$ can be represented as a Gaussian distribution with mean

\begin{equation}
\bm{K}(\bm{x_t}, \bm{x}) \left[ \bm{K}(\bm{x}, \bm{x}) + \bm{\Sigma} \right]^{-1} \bm{y}_{\lnot t}
\end{equation}
and variance
\begin{equation}
\bm{K}(\bm{x_t},\bm{x_t}) - \bm{K}(\bm{x_t}, \bm{x})^T \left[ \bm{K}(\bm{x}, \bm{x}) + \bm{\Sigma} \right]^{-1} \bm{K}(\bm{x},\bm{x_t})
\end{equation}

This means that for each $z_t$ we can analytically compute $p(z_t | \bm{K}, \bm{y}, \bm{z}_{\lnot t}, \bm{\theta})$ as the convolution of two Gaussian distributions. This concludes the resampling of $z_t$ step.

NOTE: we need to say something about Neal Algorithm 8 for sampling from new noise clusters.

@@

\begin{eqnarray}
p(\bm{\theta} | \bm{K}, \bm{y}, \bm{z}) & \propto & \int p(\bm{y} | \bm{f}, \bm{z}, \bm{\theta}) p(\bm{f} | \bm{K}) d \bm{f}, \\
& = & \int \left( \prod_t p(y_t | f_t, z_t, \bm{\theta}) \right) p(\bm{f} | \bm{K}) d \bm{f}, \\
& = & \int \left( \prod_t \mathcal{N}(y_t; f_t + \mu_{z_t}, \sigma^2_{z_t}) \right) p(\bm{f} | \bm{K}) d \bm{f}
\end{eqnarray}
NOTE: Can we analytically compute the posterior of $\mu,\sigma$? It's learning a number of Normal-Inverse Gamma distributions where the observations are Gaussians (the GP).
